\toolnameend: A Tool for Visualizing Patterns of Spreadsheet Function Combinations

### 2016-04-26 doc/paper.tex
Studies revision. Fixes #178, fixes #177, fixes #175, fixes #173, fixes #172, fixes #171, fixes #169.

		\item  \vspace{.25cm} \includegraphics[scale=.75]{glossary-lightblue} -
		\textbf{Optional argument trees} are variations of trees that show only
		formulae with a specific number of arguments. For example, the function AND can
		take any number of arguments, up to 255. By default, the tree shows data from
		AND functions with any number of argument, but these trees, marked with a darker
		blue color, can show how AND was used when \textit{only} two arguments were passed into 
		it, or \textit{only} four, and so on. See
		Figure \ref{fig:optional} for how the data between tree types differs. \end{itemize}

@@

	\subsection{Bad Smells} \label{badsmells} Fowler's description of code smells
	\cite{fowler2009refactoring} underscored an important point of code quality:
	between elegant, efficient code and bug-crippled spaghetti, there is a spectrum
	of code designs which, by themselves, are not faulty but nevertheless suggest
	vulnerabilities in code design. Since then, researchers have applied these
	concepts to spreadsheet programming, as discussed in Section
	\ref{related-work}. While some of the smells lie beyond the scope of the tool,
	such as those characterizing inter-worksheet connections, others have
	structures which create distinctive features within the tree, leading to easy
	detection through our visualization: \par

@@

		\item \textit{Multiple Operations}: A single formula comprises numerous
		functions and operators \cite{hermans2012detecting}. As a result, the formula
		tends to be prohibitively complex. In the visualization, since each nested and
		combined function creates yet another level of depth within the tree, these
		constructions can result in long, horizontal chains of nodes, as seen in
		Figure \ref{fig:longsum}.

		\item \textit{Long Parameter List}: A function uses numerous arguments, making
		it harder to understand \cite{asavametha2012detecting}. This is not a problem
		for many functions which have a limited number of arguments that it accepts.
		However, for functions that accept arguments indefinitely, such as SUM and
		CONCATENATE, it creates tall towers of argument nodes from a single function
		node.

		\item \textit{Conditional Complexity}: A logical function, particularly IF,
		nests several other logical functions inside of it, reducing code readability
		\cite{hermans2012detecting}. These can be traced manually in the trees for the
		logical functions. IF, for example, will have several more IF function nodes
		within its second or third arguments, which in turn might have more of the
		same nested within them.

@@

	\begin{figure*}[h] \centering \includegraphics[width=\textwidth]{longsum}
		 \caption{The horizontal length of a tree may indicate
			complex or redundant design.} \label{fig:longsum} \end{figure*}

@@

	\subsection{Education} When Robillard and Deline researched common obstacles
	for programmers learning new APIs, they concluded that one of the most
	important elements of a good API was the use of code examples to demonstrate
	function use. These examples, furthermore, come in a variety of forms: small
	snippets or tutorials that show only the intended function, sample applications
	that incorporate the function in a broader context, or even production code
	that employs the function but was not initially intended to be an illustrative
	example \cite{robillard2011field}. Of these categories, \toolname aligns
	entirely with the last; when a user chooses to explore a particular function
	node, they will see examples of that formula as it was used in a working
	environment, with the specifics depending on the dataset the tool is using.
	\par

	Furthermore, Robillard and Deline's study makes the case for examples that show
	"API usage patterns involving more than one method call"
	\cite{robillard2011field}. Examples that show only a single function, they
	find, are too simple, whereas those demonstrating the interaction of functions
	will equip users for more complex tasks. \toolname is especially suited for
	this focus on combination; consider again the question in the walkthrough
	(Section \ref{sec:walkthrough}), asking which functions and operations are used
	as conditions in IF. The design of this tool emphasizes the connection, then,
	between what functions like ISNA and AND return and what IF accepts, and
	specific examples for any of these combinations are only a click away.

	%Lookup example
	For a deeper example, the LOOKUP functions are particularly notorious for their
	confusing mechanisms and, as such, have warranted some extra attention with
	regards to spreadsheet design. In short, these functions are for "when you need
	to look in a single row or column and find a value from the same position in a
	second row or
	column"\footnote{https://support.office.com/en-us/article/LOOKUP-function-446d94af-663b-451d-8251-369d5e3864cb}. Examples of each can be found in Figure \ref{fig:lookupexample}.

	\begin{figure} \centering \includegraphics[width=.5\textwidth]{lookupexample}
		 \caption{Demonstrations of LOOKUP and VLOOKUP} \label{fig:lookupexample}
	\end{figure}

	LOOKUP has two forms: vector and array. In either case, the first argument is
	the value which you are trying to match in another column. If the formula is in
	the vector form, that column is the second argument, so in Figure
	\ref{fig:lookupexample}, the formula in cell D2 (\textit{=LOOKUP(3, A2:A6,
		B2:B6)}) will search for the value of 3 between cells A2 and A6. The value the
	formula returns then comes from the corresponding cell in the optional third
	argument's column; since the lookup values matches the value in cell A4, it
	returns the value from cell B4. The array form of the LOOKUP works the same
	way, except that the second argument is a two-dimensional table where it
	matches values in the table's first column and returns values from the
	last\footnote{ Depending on the dimensions of the lookup table, the function
		may substitute rows for columns.}.

	VLOOKUP works closely with this array form, except the user defines in its
	third argument which column from which to return values. In \textit{=VLOOKUP(3,
		A2:B6, 2)}, the argument ``2" tells to return the value from the second column,
	which is \textit{B2:B6} here. Furthermore, it also accepts a boolean value for
	an optional argument: when set to FALSE, it returns only exact matches, and
	when TRUE or left out, it approximates the match to the closest value beneath
	it. We demonstrate this in Figure \ref{fig:lookupexample} in the bottom two
	VLOOKUP functions, one returning ``Friday" and the other an error. HLOOKUP
	works identically but focuses on rows instead of columns.

	All of these variations may seem daunting to a beginner, but \toolname can
	help. Because of its design, users can distinguish between forms of the same
	function, as seen in Figure \ref{fig:lookups} where the one can explore both
	the frequency with which the spreadsheet programmers used these different forms
	-- three-argument vector form much more than the array form -- and then
	retrieve examples of every configuration in the dataset. The same goes for the
	V/HLOOKUP functions, as shown partially in Figure \ref{fig:vhlookups}.

@@

	\section{User Study} \label{sec:userstudy} During development, we conducted a
	brief, four-participant user study in order to detect conflicts between our
	design and the goals in Section \ref{goals}. We designed the study to be
	exploratory; given trees based in the Enron dataset and at least twenty
	minutes, we asked the participants to adopt the persona of a consultant
	evaluating a company's spreadsheet practice, allowing for questions like which
	functions on which the employees most depended or which sheets or employees
	presented the most anomalous or dangerous designs. Though we instructed them on
	how to use the tool and suggested that they visit one of the most populous
	trees in the set, we put aside any specific tasks or questions and let them
	choose their own paths. Though interpretation of the phrase ``spreadsheet
	practice" can certainly vary from participant to participant, we readily
	accepted this interpretive flexibility since we wanted a range of views and
	backgrounds despite the persona. We clarified, furthermore, that their
	observations could be directed either at the quality of the data conveyed by
	the tool or at the tool's design itself. \par

@@

	The participants were most vocal about their findings when they discovered, by
	purpose or accident, instances of heinous design in the dataset. Many times,
	they noted these examples shortly after opening a new tree, suggesting that the
	distinctive visual signatures as discussed in Section \ref{badsmells} was a
	salient force in their exploration. For example, when P3 encountered the
	initial design of the SUM tree, as seen in Figure \ref{fig:sum}, they readily
	diagnosed that either a spreadsheet author was "totally incompetent, or [the
	formula was] just some that's weird." Other than these visual cues, value
	errors, prefixed with \#s, also appear in the tree, which P4 said could be
	useful for tracking these problems if marked well.

>>>>

		\item  \vspace{.25cm} \includegraphics[scale=.75]{glossary-lightblue} -
		\textbf{Optional argument trees} are variations of trees that show only
		formulae with a specific number of arguments. For example, the function AND
		can take any number of arguments, up to 255. By default, the tree shows data
		from AND functions with any number of argument, but these trees, marked with a
		darker blue color, can show how AND was used when \textit{only} two arguments
		were passed into it, or \textit{only} four, and so on. See
		Figure~\ref{fig:optional} for how the data between tree types differs.
	\end{itemize}

@@

	\subsection{Bad Smells} \label{badsmells} Fowler's description of code
	smells~\cite{fowler2009refactoring} underscored an important point of code
	quality: between elegant, efficient code and bug-crippled spaghetti, there is a
	spectrum of code designs which, by themselves, are not faulty but nevertheless
	suggest vulnerabilities in code design. Since then, researchers have applied
	these concepts to spreadsheet programming, as discussed in
	Section~\ref{related-work}. While some of the smells lie beyond the scope of
	the tool, such as those characterizing inter-worksheet connections, others have
	structures which create distinctive features within the tree, leading to easy
	detection through our visualization: \par

@@

		\item \textit{Multiple Operations}: A single formula comprises numerous
		functions and operators~\cite{hermans2012detecting}. As a result, the formula
		tends to be prohibitively complex. In the visualization, since each nested and
		combined function creates yet another level of depth within the tree, these
		constructions can result in long, horizontal chains of nodes, as seen in
		Figure~\ref{fig:longsum}.

		\item \textit{Long Parameter List}: A function uses numerous arguments, making
		it harder to understand~\cite{asavametha2012detecting}. This is not a problem
		for many functions which have a limited number of arguments that it accepts.
		However, for functions that accept arguments indefinitely, such as SUM and
		CONCATENATE, it creates tall towers of argument nodes from a single function
		node.

		\item \textit{Conditional Complexity}: A logical function, particularly IF,
		nests several other logical functions inside of it, reducing code
		readability~\cite{hermans2012detecting}. These can be traced manually in the
		trees for the logical functions. IF, for example, will have several more IF
		function nodes within its second or third arguments, which in turn might have
		more of the same nested within them.

@@

	\begin{figure*}[h] \centering \includegraphics[width=\textwidth]{longsum}
		\caption{The horizontal length of a tree may indicate complex or redundant
			design} \label{fig:longsum} \end{figure*}

@@

	However, because \toolname makes no judgments beforehand about which designs to
	emphasize as good or bad, the tool can be used to find strange or suboptimal
	designs that exist outside these obvious categories. Consider the second
	argument of the ``='' operator, already shown expanded in
	Figure~\ref{fig:fullpic}: among the top five argument types for the position in
	an IF function were boolean values. This corresponds to designs in which
	conditions already in a boolean form were then checked against "=TRUE" or
	"=FALSE", such as in the formula \textit{IF(Formulas!C77=TRUE, "Yes")}, which
	appears in one of the sheets. Though this practice does not break anything in
	itself, it nevertheless represents redundant operations in formula design and
	indicates a possible misunderstanding of how the IF function works over the
	hundreds of examples of this boolean comparison that our tool can return.

	\subsection{Education} When Robillard and Deline researched common obstacles
	for programmers learning new APIs, they concluded that one of the most
	important elements of a good API was the use of code examples to demonstrate
	function use. Examples in APIs, furthermore, come in a variety of forms: small
	snippets or tutorials that show only the intended function, sample applications
	that incorporate the function in a broader context, or even production code
	that employs the function but was not initially intended to be an illustrative
	example~\cite{robillard2011field}. Of these categories, \toolname aligns
	entirely with the last; when a user chooses to explore a particular function
	node, they will see examples of that formula as it was used in a working
	environment, with the specifics depending on the dataset the tool is using.
	\par

	Furthermore, Robillard and Deline's study makes the case for examples that show
	``API usage patterns involving more than one method call"
	\cite{robillard2011field}. Examples that show only a single function, they
	find, are too simple, whereas those demonstrating the interaction of functions
	will equip users for more complex tasks. \toolname is especially suited for
	this focus on combination; consider again the question in the walkthrough
	(Section \ref{sec:walkthrough}), asking which functions and operations are used
	as conditions in IF. The design of this tool emphasizes the connection, then,
	between what functions like ISNA and AND return and what IF accepts, and
	specific examples for any of these combinations are only a click away.

	%Lookup example
	For a deeper example, the LOOKUP functions are some of the most popular in
	Excel yet come with a number of important but silent requirements for proper
	use, and, as such, they have warranted some extra attention with regards to
	spreadsheet design. In short, these functions are for "when you need to look in
	a single row or column and find a value from the same position in a second row
	or column"\footnote{https://support.office.com/en-us/article/LOOKUP-function-446d94af-663b-451d-8251-369d5e3864cb}. Examples of their use can be found in Figure \ref{fig:lookupexample}, and we will cover each type
	in more depth to better understand the visualizations for these functions.

	\begin{figure}[h] \centering \includegraphics[width=.5\textwidth]{lookupexample}
		\caption{Demonstrations of LOOKUP and VLOOKUP} \label{fig:lookupexample}
	\end{figure}

	LOOKUP has two forms: vector and array. In either case, the first argument is
	the value which you are trying to match in another column. If the formula is in
	the vector form, that column is the second argument, so in Figure~\ref{fig:lookupexample}, 
	the formula in cell D2 (\textit{=LOOKUP(3, A2:A6,
		B2:B6)}) will search for the value of 3 between cells A2 and A6. The value the
	formula returns then comes from the corresponding cell in the optional third
	argument's column; since the lookup values matches the value in cell A4, it
	returns the value from cell B4. The array form of the LOOKUP works the same
	way, except that the second argument is a two-dimensional table where it
	matches values in the table's first column and returns values from the last.
	Also, depending on the dimensions of the lookup table, the function may
	substitute rows for columns.

	VLOOKUP works closely with this array form, except the user defines in its
	third argument which column from which to return values. In \textit{=VLOOKUP(3,
		A2:B6, 2)}, the argument ``2" tells to return the value from the second column,
	which is \textit{B2:B6} here. Furthermore, it also accepts a boolean value for
	an optional argument: when set to FALSE, it returns only exact matches, and
	when TRUE or left out, it approximates the match to the closest value beneath
	it. We demonstrate this in Figure~\ref{fig:lookupexample} in the bottom two
	VLOOKUP functions, one returning ``Friday" and the other an error. HLOOKUP
	works identically but focuses on rows instead of columns.

	All of these variations may seem daunting to a beginner, but \toolname can
	help. Because of its design, users can distinguish between forms of the same
	function, as seen in Figure~\ref{fig:lookups} where the one can explore both
	the frequency with which the spreadsheet programmers used these different forms
	-- three-argument vector form much more than the array form -- and then
	retrieve examples of every configuration in the dataset. The same goes for the
	V/HLOOKUP functions, as shown partially in Figure~\ref{fig:vhlookups}.

@@

	\section{User Study} \label{sec:userstudy} During development, we conducted a
	brief, four-participant user study to detect conflicts between our
	design and the goals in Section \ref{goals}. We designed the study to be
	exploratory; given trees based in the Enron dataset and at least twenty
	minutes, we asked the participants to adopt the persona of a consultant
	evaluating a company's spreadsheet practice, allowing for questions like which
	functions on which the employees most depended or which sheets or employees
	presented the most anomalous or dangerous designs. Though we instructed them on
	how to use the tool and suggested that they visit one of the most populous
	trees in the set, we put aside any specific tasks or questions and let them
	choose their own paths. Though interpretation of the phrase ``spreadsheet
	practice" can certainly vary from participant to participant, we readily
	accepted this interpretive flexibility since we wanted a range of views and
	backgrounds despite the persona. We clarified, furthermore, that their
	observations could be directed either at the quality of the data conveyed by
	the tool or at the tool's design itself. \par

@@

	The participants were most vocal about their findings when they discovered, by
	purpose or accident, instances of heinous design in the dataset. Many times,
	they noted these examples shortly after opening a new tree, suggesting that the
	distinctive visual signatures as discussed in Section \ref{badsmells} was a
	salient force in their exploration. For example, when P3 encountered the
	initial design of the SUM tree, as seen in Figure \ref{fig:sum}, they readily
	diagnosed that either a spreadsheet author was "totally incompetent, or [the
	formula was] just something that's weird." Other than these visual cues, value
	errors, prefixed with \#s, also appear in the tree, which P4 said could be
	useful for tracking these problems if marked well.
